\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology

### 2022-06-08 jsdse/Manuscript.tex

\noindent%
{\it Keywords:}  open science, R programming language, phylogenetics, Open Tree of Life, pedagogy
\vfill

@@

%% \bigskip
%% \bigskip
%%
%% \noindent\fbox{%
%%     \parbox{\textwidth}{%
%%     \textbf{Box 1. The role of approachable research workflows in teacing reproducibility.}
%%       % Availability is defined as the fact or possibility that something can be bought,
%%       % used, or reached by someone \citep{availability2022cambridge}.
%%       % Accessibility is defined as "the fact of being able to be reached or obtained easily"
%%       % \citep{accessibility2022cambridge}
%%       % Yet, in practice, availability does not imply accessibility.
%%       Availability and accessibility are conceptually accepted as synonyms for ``the fact
%%       of being able to be reached" \citep{available-accessible2022cambridge}.
%%       However, accessibility's definition goes further and describes it as the
%%       ``fact of being able to be reached \textbf{easily}" \citep{accessibility2022cambridge}.
%%       A secondary definition of accessibility elaborates more on this: ``the quality
%%       of being \textbf{easy to understand}''.
%%
%%       What we realize from these definitions is that, in practice, something might
%%       be available but not necessarily accessible,
%%       hindering the realization of availability.
%%       To illustrate the practical difference between availability and accessibility we
%%       will use an allegory that goes like this:
%%
%%       ** Change marshmallow story to cake
%%
%%       ** Explain different types of barriers to accessibility within the example, then within science
%%
%%       ** Explain why we focus on psychological/social/pedagogical barriers
%%
%%       ** Gauging accessibility of a resource is tricky because it depends on an
%%       individual experience of what is in effect \textbf{easy} for someone.
%%       \textbf{Easy} is a subjective quality that fully depends on individual perceptions
%%       resulting from individual contexts.
%%       As a society we are moving towards validating and welcoming all human experiences.
%%       To improve accessibility levels in the natural sciences, it
%%       is important to account for the factors that are creating different experiences
%%       of what is \textbf{easy}.
%%
%%       ** Now, explain different types of barriers to accessibility in science
%%       (Jessica: Maybe say that lack of accessibility can be due to a number of
%%       factors (physical, resources, economic), but one critical factor is the use
%%       of impenetrable expert language.)
%%     }%
%% }
%% \bigskip

@@

Phylogenetics is a key discipline within evolutionary biology \citep{dobzhansky1973nothing}.
It focuses on investigating the history of shared ancestry of living and extinct
organisms using biological data,
and represents this evolutionary history with a diagram known as a phylogeny
or phylogenetic tree (because it grows through time and appears to have branches;
Figure \ref{fig:tree}).
Phylogenies provide the basis to study and understand all biological processes
in an evolutionary context \citep{dobzhansky1973nothing}.
Hence, it appears that improving reproducibility rates in phylogenetics has the
potential to positively impact research across the natural sciences.

@@

All of these resources
%% (the stored phylogenies, the taxonomic standard, the synthetic tree, and all
%% tools developed to process and summarize stored phylogenies),
are available for download and use from OpenTree free of cost to any user.
One way to access OpenTree resources is through its Graphical User
Interface (GUI; aka, a website or application that allows users to access computer
functionalities, in this case OpenTree resources, with mouse or keyboard clicks).
However, reducing as many manual steps as possible in research workflows is key
for reproducibility, as manual data manipulation scales poorly and is prone to
error \citep{bakken2019journey}.
OpenTree's resources are also programmatically available through its Application
Programming Interface services (APIs; aka, computer code that implements computer functionalities,
in this case OpenTree resources, that can be used by programmers
to build more functionalities),
which provide scalability and reproducibility \citep{opentreeAPIv3}.
However, this comes at a high cost for the user, which requires considerable more
computer programming experience and literacy to be able to successfully use APIs.
%% In recent years, OpenTree API services have been wrapped into software packages for R and Python
%% \citep{michonneau2016rotl, mctavish2021opentree}, open source, free of cost
%% programming languages that represent two of the most widely used programming languages
%% in the sciences \citep{baker2017scientific}. These packages should contribute to making
%% OpenTree resources more accessible to a wider user audience.
The \texttt{rotl} R package \citep{michonneau2016rotl} and the \texttt{opentree} Python module
\citep{mctavish2021opentree} have been developed to wrap OpenTree's API services
and make them available through the R and Python programming languages.
R and Python are open source and free of cost, and represent
two of the most widely used programming languages in the sciences today \citep{eglen2009quick,
baker2017scientific}.
As such, \texttt{rotl} and \texttt{opentree} software packages should contribute
to making OpenTree resources more accessible to a wider programming user audience.

@@

Besides avoiding formal language, and incorporating elements of pop culture, such as picture
character icons known as ``emojis``, to make the language more familiar to a
broader target audience (Figure \ref{fig:error}), we made an effort to specifically
complement the primary documentation by identifying
computational concepts that were assumed or were not explained in depth.
We vetted the tutorials through feedback from workshop participants as well as
individual users to identify such specialized concepts.

@@

We propose that documents developed with ``literate programming'' can be made more
accessible by choosing narratives that are relatable to a more general audience.
An easy way to do this in biology is choosing a charismatic taxon as model organism.
For a research group this can be the group they are studying. For the general audience,
a highly charismatic group such as dinosaurs will do the trick.
For example, when we presented our tutorial to the Amphibia Web Organization \citep{van2002amphibiaweb}
in January 2020, we tailored all examples to frogs and their allies.
% \citep{cite-aw-workshop}

@@

A practice that has become more and more widespread in programming-language pedagogical practices
is live programming.
One benefit of live programming is that typos and mistakes occur -- which normalizes
them for learners, and shows them how to solve them when they are outside the
classroom \citep{shannon2015live, nederbragt2020ten}.
When co-author McTavish was a postdoc, teaching an introductory programming workshop as a volunteer
with the Carpentries (a non-profit group that teaches foundational coding and
 data science skills to researchers worldwide) \citep{wilson2006swc, SWCwebsite},
a senior faculty member taking the workshop
complained that the typos were slowing things down and interfering with the pedagogy.
McTavish replied "the typos ARE the pedagogy".
This has become a slogan of sorts at the Carpentries, capturing the idea that embracing
and discussing mistakes is essential to teaching programming \citep{wilson2019teaching}.
Yet, working through mistakes is rarely done on written pedagogical materials.
Software documentation focuses on demonstrating usage function with examples that
work seamlessly, without errors. We argue that the opposite is needed to support
adoption of reproducible workflows and support long term independence in learners
and users performance \citep{gaspar2007restoring, steele2014error}.

In our tutorials, we apply this principle by demonstrating examples that do not work
as expected, and exemplifying ways to address them (Figure \ref{fig:error}).
For example, we identified inputs that would give
a wide range of warnings and errors. We then focus on providing explanations for
error and warning messages. We believe this supports users and learners to be less
intimidated by the messages and practice taking useful information out of it.
% This is probably part of the computer science curriculum, but it is not at all
% part of the natural sciences curriculum.

@@

Response from the community has been invaluable in gauging success of our teaching
materials.
Senior researchers often comment on the usefulness
of the tutorials for their research, as well as how they have supported students in using the
R packages more independently.

When developing our tutorials, we not only applied the principles elaborated here,
but we followed basic recommendations for successful reproducibility \citep{sandve2013ten}.
For example, we published the tutorials on a public, free license, free of cost, and free for
use and reuse repository and persistent website \citep{RopentreeTutorials}.
We also created a main version of the tutorial that is stable. Any updates to the tutorial are
then published as new versions, or tutorials for new workshops \citep{wilson2006swc, SWCwebsite}
Versions presented at workshops are a copy from the original repository, and
constitute a temporally stable snapshot of functions and workflows presented
during a workshop.
This ensures that the tutorials are available for the users to return to any time
they need it, and to be shared with other users and learners (Figure \ref{fig:schedule}).

@@

We note that making approachable research workflows has all the advantages of
reproducible research workflows.
It saves explanation/training time when analyses are run again by students and
collaborators.
It can also save research time for yourself when analyses are run again with more
data, a different dataset, a different organism or biological model;
It contributes to scientific efforts that can build off of each other.

% \citep{roland2002think}
Some universities have been incorporating reproducibility in their curriculum
\citep{uwlibraries2022, nigms2022}.
The focus of these resources has been for students to acquire and practice skills
to document their work.
The principles identified and outlined here can be used to set learning goals and
outcomes for new reproducibility syllabi.
Ultimately, the long term improvement of reproducibility rates in science will depend
on our ability to intentionally integrate best practices for achieving reproducibility
into the educational framework of future data scientists \citep{nasem2018data}.
Inclusion of reproducibility into the data acumen of undergraduate curriculum will
provide college learners and future researchers with the tools to develop
the fundamental skills needed to successfully create reproducible scientific workflows
and research products.

The principles to create tutorials described here facilitate adoption of software
and analysis workflows among researchers at different academic levels, from undergrads
to established researchers.
It will also help close the gap between students that had access to computational
resources (and computational training) from an early age and students that did not.
Late access to computational resources and training can occur due to lack of
economic resources, often occurring in households from underrepresented communities
and minorities \citep{google2016diversity, warner2021quantifying}.
It can also be due to gender-biased parental and community pressures,
in which male individuals are more often encouraged to perform activities related to computers,
while female individuals are discouraged \citep{warner2021quantifying}.
These principles can be used to improve not only reproducibility practices,
but also software adoption in the natural sciences.

>>>>

\noindent%
{\it Keywords:}  open science, R programming language, phylogenetics, Open Tree of Life, pedagogy, phylogenetics, datelife
\vfill

@@

%% \bigskip
%% \bigskip
%%
%% \noindent\fbox{%
%%     \parbox{\textwidth}{%
%%     \textbf{Box 1. The role of approachable research workflows in teaching reproducibility.}
%%       % Availability is defined as the fact or possibility that something can be bought,
%%       % used, or reached by someone \citep{availability2022cambridge}.
%%       % Accessibility is defined as "the fact of being able to be reached or obtained easily"
%%       % \citep{accessibility2022cambridge}
%%       % Yet, in practice, availability does not imply accessibility.
%%       Availability and accessibility are conceptually accepted as synonyms for ``the fact
%%       of being able to be reached" \citep{available-accessible2022cambridge}.
%%       However, accessibility's definition goes further and describes it as the
%%       ``fact of being able to be reached \textbf{easily}" \citep{accessibility2022cambridge}.
%%       A secondary definition of accessibility elaborates more on this: ``the quality
%%       of being \textbf{easy to understand}''.
%%
%%       What we realize from these definitions is that, in practice, something might
%%       be available but not necessarily accessible,
%%       hindering the realization of availability.
%%       To illustrate the practical difference between availability and accessibility we
%%       will use an allegory that goes like this:
%%
%%       ** Change marshmallow story to cake
%%
%%       ** Explain different types of barriers to accessibility within the example, then within science
%%
%%       ** Explain why we focus on psychological/social/pedagogical barriers
%%
%%       ** Gauging accessibility of a resource is tricky because it depends on an
%%       individual experience of what is in effect \textbf{easy} for someone.
%%       \textbf{Easy} is a subjective quality that fully depends on individual perceptions
%%       resulting from individual contexts.
%%       As a society we are moving towards validating and welcoming all human experiences.
%%       To improve accessibility levels in the natural sciences, it
%%       is important to account for the factors that are creating different experiences
%%       of what is \textbf{easy}.
%%
%%       ** Now, explain different types of barriers to accessibility in science
%%       (Jessica: Maybe say that lack of accessibility can be due to a number of
%%       factors (physical, resources, economic), but one critical factor is the use
%%       of impenetrable expert language.)
%%     }%
%% }
%% \bigskip

@@

Phylogenetics is a key discipline within evolutionary biology \citep{dobzhansky1973nothing}.
It focuses on investigating the history of shared ancestry of living and extinct organisms using biological data, and represents this evolutionary history with a diagram known as a phylogeny or phylogenetic tree (because it grows through time and appears to have branches; Figure \ref{fig:tree}).
Phylogenies provide the basis to study and understand all biological processes in an evolutionary context \citep{dobzhansky1973nothing}.
Hence, it follows that improving reproducibility rates in phylogenetics has the potential to positively impact research across the natural sciences.

@@

All of these resources 
%% (the stored phylogenies, the taxonomic standard, the synthetic tree, and all
%% tools developed to process and summarize stored phylogenies),
are available for download and use from OpenTree free of cost to any user.
One way to access OpenTree resources is through its Graphical User Interface (GUI; aka, a website or application that allows users to access and use functionalities with mouse or keyboard clicks).
However, reducing as many manual steps as possible in research workflows is key for reproducibility, as manual data manipulation scales poorly and is prone to error \citep{bakken2019journey}.
OpenTree's resources are also programmatically available through its Application Programming Interface services (APIs; aka, computer code that implements computer functionalities, in this case OpenTree resources, that can be used by programmers
to build more functionalities), which provide scalability and reproducibility \citep{opentreeAPIv3}.
However, this comes at a high cost for the user, which requires considerably more computer programming experience and literacy to be able to successfully use APIs.
%% In recent years, OpenTree API services have been wrapped into software packages for R and Python
%% \citep{michonneau2016rotl, mctavish2021opentree}, open source, free of cost
%% programming languages that represent two of the most widely used programming languages
%% in the sciences \citep{baker2017scientific}. These packages should contribute to making
%% OpenTree resources more accessible to a wider user audience.
The \texttt{rotl} R package \citep{michonneau2016rotl} and the \texttt{opentree} Python module \citep{mctavish2021opentree} have been developed to wrap OpenTree's API services
and make them available through the R and Python programming languages.
R and Python are open source and free of cost, and represent two of the most widely used programming languages in the sciences today \citep{eglen2009quick, baker2017scientific}.
As such, \texttt{rotl} and \texttt{opentree} software packages should contribute to making OpenTree resources more accessible to a wider programming user audience.

@@

Avoiding formal language, and incorporating elements of pop culture, such as picture character icons known as ``emojis``, make the language more familiar to a broader target audience (Figure \ref{fig:error}). We made an effort to specifically complement the primary documentation by identifying computational concepts that were assumed or were not explained in depth.
We vetted the tutorials through feedback from workshop participants as well as individual users to identify such specialized concepts.

@@

We propose that documents developed with ``literate programming'' can be made more accessible by choosing narratives that are relatable to a more general audience.
An easy way to do this in biology is choosing a charismatic taxon as a model organism.
For a research group this can be the group they are studying. For the general audience, a highly charismatic group such as dinosaurs should work well.
For example, when we presented our tutorial to the Amphibia Web Organization \citep{van2002amphibiaweb} in January 2020, we tailored all examples to frogs and their allies.
% \citep{cite-aw-workshop}

@@

An activity that has become increasingly widespread in programming-language education is live programming. During live programming, and instructor writes code and executes it in a way that is visible to all learners through a screen \citep{guzdial2013lure, selvaraj2021live}.
One benefit of this practice is that typos and mistakes occur, normalizing them for learners. 
Watching an instructor handling errors, demonstrates learners on how to solve them when they are outside theclassroom \citep{shannon2015live, nederbragt2020ten}.
When co-author McTavish was a postdoc, teaching an introductory programming workshop as a volunteer with the Carpentries (a non-profit group that teaches foundational coding and data science skills to researchers worldwide) \citep{wilson2006swc, SWCwebsite},
a senior faculty member taking the workshop complained that the typos were slowing things down and interfering with the pedagogy.
McTavish replied "the typos ARE the pedagogy".
This has become a slogan of sorts at the Carpentries, capturing the idea that embracing and discussing mistakes is essential to teaching programming \citep{wilson2019teaching}.
Yet, working through mistakes is rarely done on written pedagogical materials.
Software documentation focuses on demonstrating usage function with examples that work seamlessly, without errors. 
We argue that the opposite is needed to support adoption of reproducible workflows and support long term independence in learners and users performance \citep{gaspar2007restoring, steele2014error}.

In our tutorials, we apply this principle by demonstrating examples that do not work as expected, and exemplifying ways to address them (Figure \ref{fig:error}).
For example, we identified inputs that would give a wide range of warnings and errors. We then focus on providing explanations for these errors and warning messages. We believe this supports users and learners to be less intimidated by the messages, and to practice taking useful information out of them.
% This is probably part of the computer science curriculum, but it is not at all
% part of the natural sciences curriculum.

@@

Response from the community has been invaluable in gauging success of our teaching materials.
Senior researchers often comment on the usefulness of the tutorials for their research, as well as how they have supported students in using the demosntrated R packages more independently.

When developing our tutorials, we not only applied the principles elaborated here,
but we followed basic recommendations for successful reproducibility \citep{sandve2013ten}.
For example, we published the tutorials on a persistent and public website \citep{RopentreeTutorials} that adheres to the four r's of openness, by being free license, free of cost, and thus free for use, reuse, redistribute, revise and remix \citep{hilton2010four}. 
To make the website persistent, any updates to the tutorial are published as new versions.
Versions presented at workshops are a copy from the original repository, and constitute a temporally stable snapshot of functions and workflows presented during a live workshop \citep{wilson2006swc, SWCwebsite}.
This ensures that the tutorials are available for the users to return to any time they need it, and to be shared with other users and learners (Figure \ref{fig:schedule}).
This approach not only teaches reproducibility, but also makes the teaching process itself reproducibile \citep{dogucu_tools_2022}.

@@

We note that making approachable research workflows has all the advantages of reproducible research workflows.
It saves time during explanation and training, when analyses are run by new collaborators and students.
It can also save research time for yourself, when analyses are run again with more data, a different dataset, a different organism or biological model.
It contributes to scientific efforts that can build off of each other.

% \citep{roland2002think}
Some universities and academic groups have started to incorporate reproducibility as a subject into their curriculum
\citep{uwlibraries2022, nigms2022, psyteachr2022}.
The focus of these resources has been for students to acquire and practice skills to document their work.
The principles identified and outlined here can be used to set learning goals and outcomes for new reproducibility syllabi.
Ultimately, the long term improvement of reproducibility rates in science will depend on our ability to intentionally integrate best practices for achieving reproducibility into the educational framework of future data scientists \citep{nasem2018data}.
Inclusion of reproducibility into the data acumen of undergraduate curriculum will provide college learners and future researchers with the tools to develop the fundamental skills needed to successfully create reproducible scientific workflows and research products.

The principles to create tutorials described here facilitate adoption of software and analysis workflows among researchers at different academic levels, from undergrads to established researchers.
It will also help close the gap between students that had access to computational resources (and computational training) from an early age and students that did not.
Late access to computational resources and training can occur due to lack of economic resources, often occurring in households from underrepresented communities and minorities \citep{google2016diversity, warner2021quantifying}.
It can also be due to gender-biased parental and community pressures, in which male identifying individuals are more often encouraged to perform activities related to computers, while female identifying individuals are discouraged, starting from as early as elementray school \citep{master_gender_2021}.
These principles can be used to improve not only reproducibility practices, but also software adoption in the natural sciences.
